---
title: "HW2"
author: "Jiaheng Cai"
date: "10/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival) 
library(KMsurv)
```

## 3.6

### (a)

$$\begin{align*}
L&\propto\prod_{i=1}^{n}f(T_i)^{\delta_i}S(T_i)^{1-\delta_i}\\
&=\prod_{i=1}^n(\lambda e^{-\lambda T_i})^{\delta_i}(e^{-\lambda T_i})^{1-\delta_i}\\
&=\lambda^{\sum_{i=1}^n\delta_i}e^{-\lambda\sum_{i=1}^nT_i}\\
&=\lambda^6e^{-180\lambda}
\end{align*}$$

### (b)

$$\begin{align*}
L&\propto\prod_{i=1}^{n}f(T_i)^{\delta_i}S(T_i)^{1-\delta_i}\\
&=\prod_{i=1}^n(\alpha\theta T_i^{\alpha -1}e^{-\theta T_i^\alpha})^{\delta_i}(e^{-\theta T_i^\alpha})^{1-\delta_i}\\
&=(\alpha\theta)^4\prod_{i=1}^n
(T_i^{\alpha -1})^{\delta_i}e^{-\theta \sum_{i=1}^n(T_i^\alpha)}\\
&=(\alpha\theta)^4 (89100)^{\alpha -1}e^{-\theta (11^\alpha+12^\alpha+15^\alpha+33^\alpha+45^\alpha+28^\alpha+16^\alpha+17^\alpha+19^\alpha+30^\alpha)}
\end{align*}$$

### (c)

$$\begin{align*}
L&=\prod_{i=1}^n\frac{f(x_i^*)^{\delta_i}S(x_i^*)^{1-\delta_i}}{S(y_i^*)}\\
&=\prod_{i=1}^n\frac{(\alpha\theta x_i^{\alpha -1})^{\delta_i}(e^{-\theta x_i^\alpha})^{\delta_i}(e^{-\theta x_i^\alpha})^{1-\delta_i}}{e^{-\theta y_i ^\alpha}}\\
&=\frac{(\alpha \theta)^4 89100^{\alpha -1}e^{-\theta (11^\alpha+12^\alpha+15^\alpha+33^\alpha+45^\alpha+28^\alpha)}}{e^{-\theta(5^\alpha+8^\alpha+12^\alpha+24^\alpha+32^\alpha+17^\alpha)}}
\end{align*}$$

## 12.1

### (a)

```{r}
data(tongue, package= "KMsurv")
tongue1 = tongue[tongue$type == 1,]
tongue2 = tongue[tongue$type == 2,]

fit1 <- survreg(Surv(time, delta) ~ 1, data = tongue1, dist = "weibull")
fit2 <- survreg(Surv(time, delta) ~ 1, data = tongue2, dist = "weibull")

fit1e <- survreg(Surv(time, delta) ~ 1, data = tongue1, dist = "exponential")
fit2e <- survreg(Surv(time, delta) ~ 1, data = tongue2, dist = "exponential")
```

lambdas and alphas
```{r}
mu1    <- fit1$coefficients[1]
sigma1 <- fit1$scale

(lambda1   <- exp(-mu1 / sigma1))
(alpha1    <- 1 / sigma1)

mu2    <- fit2$coefficients[1]
sigma2 <- fit2$scale

(lambda2   <- exp(-mu2 / sigma2))
(alpha2    <- 1 / sigma2)
```

se lambda1, se alpha1, se lambda2, se alpha2
```{r}
varlm1 = exp(-2*mu1/sigma1)*(fit1[["var"]][1,1]/sigma1^2 + mu1^2*fit1[["var"]][2,2]/sigma1^2 -2*mu1*sigma1*fit1[["var"]][1,2]/sigma1^3)
(sqrt(varlm1))

(sealp1 = sqrt(fit1[["var"]][2,2]/sigma1^2))

varlm2 = exp(-2*mu2/sigma2)*(fit2[["var"]][1,1]/sigma1^2 + mu2^2*fit2[["var"]][2,2]/sigma2^2 -2*mu2*sigma2*fit2[["var"]][1,2]/sigma2^3)
(sqrt(varlm2))

(sealp2 = sqrt(fit2[["var"]][2,2]/sigma2^2))
```

### (b)

```{r}
sqetw1 = ((alpha1 - 1)/sealp1)^2
pchisq(sqetw1, 1, lower.tail = FALSE)

sqetw2 = ((alpha2 - 1)/sealp2)^2
pchisq(sqetw2, 1, lower.tail = FALSE)

(pchisq(2*(fit1$loglik[1] - fit1e$loglik[1]), 1, lower.tail = FALSE))
(pchisq(2*(fit2$loglik[1] - fit2e$loglik[1]), 1, lower.tail = FALSE))
```
### (c)

$$\begin{align*}
S(y)&=0.5\\
e^{-e^{ \frac{(y- \mu ) }{ \sigma }}}&=0.5\\
-e^{ \frac{(y- \mu ) }{ \sigma }} &=\log(0.5)\\
\frac{(y- \mu ) }{ \sigma }&=\log(-\log(0.5))\\
y&=\log(-\log(0.5))\sigma +\mu
\end{align*}$$

```{r}
(exp(log(-log(0.5))*sigma1+mu1))
(exp(log(-log(0.5))*sigma2+mu2))
```

By delta method:

$$\sqrt{n}(g\begin{pmatrix}\widehat{\mu}\\\widehat{\sigma}\end{pmatrix}-g\begin{pmatrix}\mu_i\\\sigma_i\end{pmatrix})\stackrel{D}{\longrightarrow}N(0,g'\begin{pmatrix}\mu_i\\\sigma_i\end{pmatrix}\Sigma g'\begin{pmatrix}\mu_i\\\sigma_i\end{pmatrix}^T)$$
```{r}
gd1 = matrix(0,1,2)
gd1[1,1] = exp(log(-log(0.5))*sigma1+mu1)
gd1[1,2] = log(-log(0.5))*exp(log(-log(0.5))*sigma1+mu1)
vcov1 = fit1$var
vcov1[1,2] = vcov1[1,2]*sigma1
vcov1[2,1] = vcov1[2,1]*sigma1
vcov1[2,2] = vcov1[2,2]*sigma1^2

sqrt(gd1%*%vcov1%*%t(gd1))

gd2 = matrix(0,1,2)
gd2[1,1] = exp(log(-log(0.5))*sigma2+mu2)
gd2[1,2] = log(-log(0.5))*exp(log(-log(0.5))*sigma2+mu2)
vcov2 = fit2$var
vcov2[1,2] = vcov2[1,2]*sigma2
vcov2[2,1] = vcov2[2,1]*sigma2
vcov2[2,2] = vcov2[2,2]*sigma2^2

sqrt(gd2%*%vcov2%*%t(gd2))
```


### (d)
```{r}
fita <- survreg(Surv(time, delta) ~ type, data = tongue, dist = "weibull")
summary(fita)
mua    <- fita$coefficients[1]
betaa <- fita$coefficients[2]
sigmaa <- fita$scale

lambdaa   <- exp(-mua / sigmaa)
alphaa    <- 1 / sigmaa
```
fail to reject h0 : type = 0 for both test

RR is
```{r}
exp(betaa/sigmaa)
```
Again, by delta method:
$$\sqrt{n}(g\begin{pmatrix}\widehat{\beta}\\\widehat{\sigma} \end{pmatrix}-g\begin{pmatrix}\beta_i\\\sigma_i\end{pmatrix})\stackrel{D}{\longrightarrow}N(0,g'\begin{pmatrix}\beta_i\\\sigma_i\end{pmatrix}\Sigma g'\begin{pmatrix}\beta_i\\\sigma_i\end{pmatrix}^T)$$
```{r}
gdra = matrix(0,1,2)
gdra[1,1] = 1/sigmaa
gdra[1,2] = -betaa/sigmaa^2
vcovra = fita$var[-1,-1]
vcovra[1,2] = vcovra[1,2]*sigmaa
vcovra[2,1] = vcovra[2,1]*sigmaa
vcovra[2,2] = vcovra[2,2]*sigmaa^2

sqrt(gdra%*%vcovra%*%t(gdra))
#CI low
(exp(betaa/sigmaa-1.96*sqrt(gdra%*%vcovra%*%t(gdra))))
#CI up
(exp(betaa/sigmaa+1.96*sqrt(gdra%*%vcovra%*%t(gdra))))
```

AF is
```{r}
exp(betaa)
```
and CI is
```{r}
#CI low
exp(betaa-1.96*0.351)
#CI up
exp(betaa+1.96*0.351)
```
The median lifetime for diploid patients is between 0.26 and 1.02 times that if aneuploid

## 12.2

### (a)

```{r}
data("kidney", package = "KMsurv")
kidney1 = kidney[kidney$type == 1,]
kidney2 = kidney[kidney$type == 2,]

fitk1 <- survreg(Surv(time, delta) ~ 1, data = kidney1, dist = "weibull")
fitk2 <- survreg(Surv(time, delta) ~ 1, data = kidney2, dist = "weibull")

fitk1e <- survreg(Surv(time, delta) ~ 1, data = kidney1, dist = "exponential")
fitk2e <- survreg(Surv(time, delta) ~ 1, data = kidney2, dist = "exponential")
```

```{r}
muk1    <- fitk1$coefficients[1]
sigmak1 <- fitk1$scale

(lambdak1   <- exp(-muk1 / sigmak1))
(alphak1    <- 1 / sigmak1)

muk2    <- fitk2$coefficients[1]
sigmak2 <- fitk2$scale

(lambdak2   <- exp(-muk2 / sigmak2))
(alphak2    <- 1 / sigmak2)
```

se lambda1, se alpha1, se lambda2, se alpha2
```{r}
varlmk1 = exp(-2*muk1/sigmak1)*(fitk1[["var"]][1,1]/sigmak1^2 + muk1^2*fitk1[["var"]][2,2]/sigmak1^2 -2*muk1*sigmak1*fitk1[["var"]][1,2]/sigmak1^3)
(sqrt(varlmk1))

(sealpk1 = sqrt(fitk1[["var"]][2,2]/sigmak1^2))

varlmk2 = exp(-2*muk2/sigmak2)*(fitk2[["var"]][1,1]/sigmak2^2 + muk2^2*fitk2[["var"]][2,2]/sigmak2^2 -2*muk2*sigmak2*fitk2[["var"]][1,2]/sigmak2^3)
(sqrt(varlmk2))

(sealpk2 = sqrt(fitk2[["var"]][2,2]/sigmak2^2))
```

### (b)

```{r}
sqetwk1 = ((alphak1 - 1)/sealpk1)^2
pchisq(sqetwk1, 1, lower.tail = FALSE)

sqetwk2 = ((alphak2 - 1)/sealpk2)^2
pchisq(sqetwk2, 1, lower.tail = FALSE)

(pchisq(2*(fitk1$loglik[1] - fitk1e$loglik[1]), 1, lower.tail = FALSE))
(pchisq(2*(fitk2$loglik[1] - fitk2e$loglik[1]), 1, lower.tail = FALSE))
```

### (d)
```{r}
fitka <- survreg(Surv(time, delta) ~ type, data = kidney, dist = "weibull")
summary(fitka)
muka    <- fitka$coefficients[1]
betaka <- fitka$coefficients[2]
sigmaka <- fitka$scale

lambdaka   <- exp(-muka / sigmaka)
alphaka    <- 1 / sigmaka
```
fail to reject h0:type = 0 for both test

RR is
```{r}
exp(betaka/sigmaka)
```


```{r}
gdrka = matrix(0,1,2)
gdrka[1,1] = 1/sigmaka
gdrka[1,2] = -betaka/sigmaka^2
vcovrka = fitka$var[-1,-1]
vcovrka[1,2] = vcovrka[1,2]*sigmaka
vcovrka[2,1] = vcovrka[2,1]*sigmaka
vcovrka[2,2] = vcovrka[2,2]*sigmaka^2

sqrt(gdrka%*%vcovrka%*%t(gdrka))
(exp(betaka/sigmaka-1.96*sqrt(gdrka%*%vcovrka%*%t(gdrka))))
(exp(betaka/sigmaka+1.96*sqrt(gdrka%*%vcovrka%*%t(gdrka))))
```

AF is
```{r}
exp(betaka)
```
and CI is
```{r}
exp(betaka-1.96*0.469)
exp(betaka+1.96*0.469)
```
The median lifetime for surgical patients is between 0.26 and 1.02 times that if percutaneous